Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-25 trunk/paper/drinking.tex

Automated reasoning in the Web Ontology Language (OWL) can be employed
to automatically compute the generalization hierarchy underlying an
ontology as well as for verification of data consistency and complex
queries \cite{Hoehndorf2011incon, Hoehndorf2011models}.

Layered approach: although OWL has high expressivity, some things
cannot be expressed in OWL; some things should not be expressed in OWL
for specific applications due to tractability reasons

Most applications rely on graph structure of ontology...

@@

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while a particular {\em goal line}
can participate in the same process in a different role.

@@

\subsection{Gene Ontology}
quick intro...

mention regulation relations...

regulation of X processes

\subsection{PATO and the EQ model}

@@

\subsection{Cellular Phenotype Ontology}

A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

@@

\begin{table*}
  \begin{tabular}{l|l|l|l}
    & Increased blood flow rate & Normal blood flow rate & Decreased
    blood flow rate \\
    Increased frequency &increased&increased&?\\
    Normal frequency &increased&normal&decreased\\
    Decreased frequency &?&decreased&decreased
  \end{tabular}
  \caption{\label{...}Total blood flow rate...}
\end{table*}

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can be increased or
decreased not within a single {\em cytoplasmic streaming} process but
rather the total flow rate, as a summation over all cytoplasmic
streaming processes that occur within an organism (or a particular
anatomical location), is increased or decreased.

@@

>>>>

Automated reasoning in the Web Ontology Language (OWL) can be employed
to automatically compute the generalization hierarchy underlying an
ontology as well as for verification of data consistency and complex
queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
efficient automated reasoners are available to process OWL ontologies
\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
to support even large ontologies by further reducing the expressivity
of OWL in order to enable polynomial-time inferences. In particular
the OWL EL profile was found to provide the expressivity required for
most biomedical ontologies \cite{el4, elvira}, and highly optimized
OWL EL reasoners are available or under development to support
reasoning over very large ontologies \cite{el4, cbreasoner}.

A high expressivity is required to accurately specify complex axioms
that constrain the domain under investigation, and languages with
higher expressivity than OWL are often required in the biomedical
domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
the other hand, automated reasoning over large ontologies and
associated datasets benefits from languages with a low complexity of
inferences in which complex axioms cannot be formulated. Therefore, a
possible solution is to use a layered approach: to specify the meaning
of terms using an expressive language, and derive the axioms that
must obtain in a weaker language using deductive inferences.

@@

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while another person can participate
in the same process but in a different role (such as {\em Referee}).

@@

\subsection{Gene Ontology}
The Gene Ontology provides a set of ontologies for molecular and
cellular biology, originally designed to support structured
annotations for genes and gene products in all species with respect to
molecular function (MF), biological process (BP) and cellular
component (CC). MF and BP both describe processes, but at different
spatiotemporal scales; in particular, BP includes processes that
unfold within cells and within tissues and organs of multicellular
organisms. Gene product annotations identify participants in the
processes.

Over time, GO development has increasingly emphasized a normalized
approach that includes supplementing existing human-readable text
description with formally specified explicit definitions for GO
classes. The formalization of GO is readily apparent in its
representation of biological regulation.

Regulatory processes may regulate other processes, at either the MF or
BP scale, or biological qualities. GO accordingly includes three broad
categories of regulation terms, regulation of molecular function,
regulation of biological process, and regulation of biological
quality. The first two are explicitly defined entirely with respect to
other GO terms, whereas the third comprises classes in which the
regulated qualities are specified by terms from PATO (see below) or
anatomy ontologies.

All GO regulation terms use one of three relations, {\bf regulates},
{\bf negatively\_regulates} and {\bf positively\_regulates}, to link
regulation terms to process or quality terms. The {\bf regulates}
relations are defined in terms of qualities: a regulatory process
causes a change in magnitude to some quality, which in turn has an
effect on the frequency, rate or duration of some other type of
process. Effects that results in increases and decreases use {\bf
  positively\_regulates} and {\bf negatively\_regulates} respectively
\cite{Mungall2010go}. The existing ontology structure would also
support the addition of subclasses to distinguish, for example,
regulation of the rate of a process from regulation of its duration or
time of onset.

\subsection{PATO and the EQ model}
The Phenotype And Trait Ontology (PATO) was envisaged and designed to
provide a platform for allowing the integration of quantitative and
qualitative phenotype related information across different levels of
granularity of domain knowledge granularity and species
\cite{Gkoutos2005}.  PATO forms an ontology of phenotype qualities
that form basic entities that we can perceive and/or measure such as
colors, sizes, rates etc. One of it's classification axis lies on the
basis of the basic type of entity these qualities inhere to and
distinguishes between qualities of physical objects and qualities of
processes.

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005a}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for the {\tt MP:0009007} term from
the Mammalian Phenotype) it could be used to provide an equivalence
statement between that class and the above PATO-based description.

@@

\subsection{Cellular Phenotype Ontology}
A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

@@

\begin{table*}
  \centering
  \begin{tabular}{l|l|l|l}
    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
    cytoplasmic flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Total cytoplasmic flow rate...}
\end{table*}

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can be increased or
decreased not within a single {\em cytoplasmic streaming} process but
rather the total cytoplasmic flow rate, as a summation over all
cytoplasmic streaming processes that occur within an organism (or a
particular anatomical location), is increased or decreased. While a
flow rate of a single {\em cytoplasmic streaming} process is a quality
of that process, an increased {\em total} cytoplasmic flow rate is a
quality of the regulation of {\em cytoplasmic streaming}. In
particular, it is possible for an organism to have a normal --- or
even a decreased --- cytoplasmic flow rate in each individual
cytoplasmic streaming process while at the same time having an
increased total cytoplasmic flow rate due to a large increase in the
frequency of occurrence of cytoplasmic streaming processes. Similarly,
the frequency of occurrence of cytoplasmic streaming may be normal or
decreased while the total cytoplasmic flow rate is increased due to an
increased cytoplasmic flow rate in each individual cytoplasmic
streaming process. Table \ref{tbl:flow} illustrates the dependencies
between rates of individual processes, their frequency of occurrence
and the total rate of these processes. We include total rates as
regulatory abnormalities in the CPO since these are the attributes of
processes that are often measured or observed, while the rates of
individual processes are inferred following a schema such as Table
\ref{tbl:flow}.

@@

When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
XX classes...
Classifies in...
